\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-09 coqimpl.tex
Elaborate Compilation Process. Example for prec_discriminate.

% \begin{figure}[!htb]\YZ{Change this one into STLC example, to see the layout}
%   \begin{minipage}[t]{0.32\linewidth}
% \begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
% Family B.
%   FInductive b : Set 
%     := tt : b | ff : b.
%   Family neg_handler. 
%     Field tt : self_B.b 
%       := self_B.ff.
%     Field ff : self_B.b 
%       := self_B.tt.
%   EndFamily.
%   FRecursor neg 
%     about self_B.b 
%     motive (fun _ => self_B.b)
%     using self_B.neg_handler
%     by _rec.
%   Field example := 
%     self_B.neg self_B.tt. 
% EndFamily.
% Family B2 extends B.
%   Extend FInductive b : Set 
%     := uu : b.
%   Extend Family neg_handler.
%     Field uu : self_B2.b 
%       := self_B2.uu.
%   EndFamily. 
% EndFamily.
% \end{lstlisting}
%   \end{minipage}
%   \begin{minipage}[t]{0.65\linewidth}
% \begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}
%   (* Abstraction for Inductive Type *)
% Module Type b_3 (self_B: EmptySig_4).
% Parameter (b : Set). Parameter (tt ff : b). End b_3.      
% Module Type b_6 := b_3 EmptyMod.
%   (* Recursor to Set for b *)
% Module b_rec_12 (self_B: b_6).
% Definition __recursor_type_b_rec :=
%   forall P : self_B.b -> Set,
%   P self_B.tt -> P self_B.ff 
%   -> forall __i : self_B.b, P __i. End b_rec_12. 
%   (* Field B.neg_handler.ff *)
% Module ff_24 (self_B: b_6) 
%              (self_neg_handler: tt_23 self_B).
% Definition ff : self_B.b := self_B.tt. End ff_24.
%   (* Intermediate Module solely for type checking *)
% Module v_33_34 (self_B: neg_handler_26).
% Include b_rec_12 self_B.
% Parameter (recursor_for_type_checking : __recursor_type_b_rec).
% Definition term_for_type_checking :=
%   recursor_for_type_checking (fun _ : self_B.b => self_B.b)
%     self_B.neg_handler.tt
%     self_B.neg_handler.ff.  
% End v_33_34.
% \end{minted}
%   \end{minipage}
% \caption{Example Code for Inductive Type}\label{fig:plugin-example2}
% \end{figure}

\begin{figure}[!htb]
  \begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
(* ... *)
  Family subst_internal. 
  Final Field tm_var : ...
:= fun s x t => 
    if (eqb x s) 
    then t else (tm_var s).
  Final Field tm_abs : ...
:= fun s (b : tm)   
   (recb : id → tm → tm) 
   x t => 
    if (eqb x s) 
    then (tm_abs s body)
    else (tm_abs s (recb x t)).
(* ... *)
  EndFamily.
  FRecursor subst 
    about tm 
    motive (fun _ => id → tm → tm)
    using subst_internal
    by _rec.
(*
Field test := (subst (tm_var 0) 0 (tm_var 0)).@\YZ{Why is "test" commented out?}\EDJreply{Because I want to show I can do it but I don't really want to d -- if I really do it the compilation result shown later will need to have this information.}@
*)
(* ... *)
EndFamily.
Family STLC_bool extends STLC.
(* ... *)
Extend FInductive tm : Set :=
  | tm_true : tm ...

Extend Family subst_internal.
Final Field tm_true : ...
  := fun x t => tm_true.
(* ... *)
EndFamily. 
Inherits subst.
(* ... *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.65\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}

Module Type STLC_409.
  Parameter (tm : Set).
  Parameter (tm_var : id → tm).
  Parameter (tm_abs : id → tm → tm).
  (* ... and all the fields defined before
  and the abstraction for inductive type ...*)
End STLC_409.
(* Collect eliminator to Set for tm *)
Module tm_rec_425 (STLC: STLC_409).
  Definition __recursor_type_tm_rec :=
    forall P : STLC.tm -> Set,
    (forall n : id, P (STLC.tm_var n)) ->
    (forall (n : id) (i : STLC.tm),
    P i -> P (STLC.tm_abs n i)) ->
    ... -> forall i : STLC.tm, P i.
End tm_rec_425.
(* Compiled Field STLC.subst_internal.tm_var *)
Module tm_var_411 (STLC: STLC_409)
  (subst_internal: EmptySig).
  Definition tm_var :
    forall (s : id) (x : id) (t : STLC.tm), STLC.tm :=
    fun s x t ⇒ if eqb x s then t else STLC.var s.
End tm_var_411. (* ... and more *)
(* Aggregate into one subst_internal*)
Module subst_internal_410 (STLC: STLC_409). 
  Include tm_var_411 STLC. (* And other cases ... *)
End subst_internal_410.
Module v_33_34 (STLC: STLC_409).
(* Intermediate Module solely for type checking *)
  Include tm_rec_425 STLC.
  Parameter (recursor_for_type_checking : __recursor_type_tm_rec).
  Definition term_for_type_checking :=
    recursor_for_type_checking (fun _  ⇒ id → STLC.tm → STLC.tm)
      subst_internal_410.tm_var
      subst_internal_410.tm_abs ... .
End v_33_34.@\YZ{It'd be useful to (1) show how the parts of the STLC family are assembled and recursor inserted, and (2) show the code compiled for STLC\_bool and how code from STLC is shared with STLC\_bool.}@
\end{minted}
  \end{minipage}
\caption{Exemplar STLC Code, especially about Inductive Type}\label{fig:plugin-example2}
\end{figure}

@@

Let's look at the exemplar \cref{fig:plugin-example2}, the example of
"subst" on STLC. When we define the extensible inductive type "tm", all
the following fields (e.g., "subst_internal.tm_var") can only know "tm"
is a type (of "Set") and there are at least two constructors for it,
just like what we declare in ``the abstracted interface''---the compiled
module type "STLC_409". We cannot export
%\YZ{export or put?}\EDJreply{Ok.
%I just figure out how my wording is confusing. put is absolutely correct
%but by put I want to mean ``export''. I think I use ``export'' in
%several places...}
the eliminator of the inductive type "tm" to "STLC_409".
Otherwise all the subsequent fields (e.g. "subst_internal.tm_var" and
the corresponding "tm_var_411") would not be able to be inherited to the
context where "tm" is extended with a third constructor, \textbf{since
the mere existence of the recursor will assert that there are only two
constructors} in "STLC_bool", not \textbf{extensible} at all!

However, we later still need the concrete eliminator on "tm" for doing
exhaustiveness checking (as in \ref{chg:extensible-inductive-type}),
thus we will store the corresponding info somewhere.  Here we show one
exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module
"tm_rec_425". This concrete eliminator will be used by our "FRecursor".
For example, in \cref{fig:plugin-example2}, "FRecursor" command will
feed Coq a module "v_33_34" for doing type-checking with specified
handlers (e.g. "neg_handler") during compilation---what it does is
basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq.\YZ{
  Sounds like we can still claim the plugin is doing type checking.
  It's just that the checking is done by generating some Coq code and
  then having Coq check it.
}\EDJreply{Yes! If my phrasing doesn't make things clear (i.e. our plugin is doing typechecking),
feel free to change the phrasing for me}
Note that, at this point of the example,
"FRecursor" will only do type check. The compilation of "FRecursor" into
a component of the compiled module only happens when closing the whole
family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and
the (vanilla) inductive type in that compiled module. Thus a "FRecursor"
field is unlike other fields, even though it is exposed as a plain
function for the succeeding fields in the programming interface.

@@

Importantly, the recursor field (e.g. "subst") is not actually
inheritable, because it is not compiled like any other fields.
Recall how other fields are compiled into reusable pieces of
(parametrized) modules; a recursor field is not compiled like that.
For all the children family, we have to reconstruct and override the
"FRecursor" field with a ``new'' recursor using (maybe the same) handler
family.  This is affordable since each aggregated recursor handler
doesn't need to be rechecked---look at "STLC_bool.neg_handler", there
are only  new handlers for "tm_true" and the handlers (mundane fields)
from the parent "B.subst_internal.tm_var" have already been type-checked
and inherited (since they only rely on ``the abstracted interface''
module "STLC_409"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "tm_var_411"---doesn't require another type-check and is
re-used during the type-checking and compilation of "STLC_bool.subst".
For example, for the new "subst", when compiling family "STLC_bool" into
a module, at the appropriate place our plugin will insert
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and the (vanilla
but extended and modified) inductive type in that compiled module.\YZ{Where is this "tm" inductive type in figure 6?} Note
that, even though "tm_rec" and "tm" has the same name as before, they
are in the newly-compiled module so they refer to totally different
things.

@@

Though partial recursor is provable by simply using "FRecursor", for convenience, we support using "FScheme" to automatic generate it. At the meantime, we implement a tactic "prec_discriminate" that when fed with the appropriate partial recursor, it can discriminate an absurd equation between two different constructors of extensible inductive type.

>>>>

% \begin{figure}[!htb]\YZ{Change this one into STLC example, to see the layout}
%   \begin{minipage}[t]{0.32\linewidth}
% \begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
% Family B.
%   FInductive b : Set 
%     := tt : b | ff : b.
%   Family neg_handler. 
%     Field tt : self_B.b 
%       := self_B.ff.
%     Field ff : self_B.b 
%       := self_B.tt.
%   EndFamily.
%   FRecursor neg 
%     about self_B.b 
%     motive (fun _ => self_B.b)
%     using self_B.neg_handler
%     by _rec.
%   Field example := 
%     self_B.neg self_B.tt. 
% EndFamily.
% Family B2 extends B.
%   Extend FInductive b : Set 
%     := uu : b.
%   Extend Family neg_handler.
%     Field uu : self_B2.b 
%       := self_B2.uu.
%   EndFamily. 
% EndFamily.
% \end{lstlisting}
%   \end{minipage}
%   \begin{minipage}[t]{0.65\linewidth}
% \begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}
%   (* Abstraction for Inductive Type *)
% Module Type b_3 (self_B: EmptySig_4).
% Parameter (b : Set). Parameter (tt ff : b). End b_3.      
% Module Type b_6 := b_3 EmptyMod.
%   (* Recursor to Set for b *)
% Module b_rec_12 (self_B: b_6).
% Definition __recursor_type_b_rec :=
%   forall P : self_B.b -> Set,
%   P self_B.tt -> P self_B.ff 
%   -> forall __i : self_B.b, P __i. End b_rec_12. 
%   (* Field B.neg_handler.ff *)
% Module ff_24 (self_B: b_6) 
%              (self_neg_handler: tt_23 self_B).
% Definition ff : self_B.b := self_B.tt. End ff_24.
%   (* Intermediate Module solely for type checking *)
% Module do_tc (self_B: neg_handler_26).
% Include b_rec_12 self_B.
% Parameter (recursor_for_type_checking : __recursor_type_b_rec).
% Definition term_for_type_checking :=
%   recursor_for_type_checking (fun _ : self_B.b => self_B.b)
%     self_B.neg_handler.tt
%     self_B.neg_handler.ff.  
% End do_tc.
% \end{minted}
%   \end{minipage}
% \caption{Example Code for Inductive Type}\label{fig:plugin-example2}
% \end{figure}

\begin{figure}[!htb]
  \begin{minipage}[t]{0.32\linewidth}
\begin{lstlisting}[language=Coq,  escapeinside={@}{@}]
Family STLC.
  FInductive tm : Set := ... 
  FScheme tm_prec PRecT about self__STLC.tm.
  FScheme Comp about self__STLC.tm_prec.
  (* Generate Partial Recursor 
  and the computational axiom *)
  Family subst_internal. 
  Final Field tm_var : ...
:= fun s x t => 
    if (eqb x s) 
    then t else (tm_var s).
  Final Field tm_abs : ...
:= fun s (b : tm)   
   (recb : id → tm → tm) 
   x t => 
    if (eqb x s) 
    then (tm_abs s body)
    else (tm_abs s (recb x t)).
(* ... *)
  EndFamily.
  FRecursor subst 
    about tm 
    motive (fun _ => id → tm → tm)
    using subst_internal
    by _rec.
(* Field test := (subst (tm_var 0) 0 (tm_var 0)).@\YZ{Why is "test" commented out?}\EDJreply{Because I want to show I can do it but I don't really want to d -- if I really do it the compilation result shown later will need to have this information.}@ *)
(* ... *)
EndFamily.
Family STLC_bool extends STLC.
(* ... *)
Extend FInductive tm : Set :=
  | tm_true : tm ...
Extend Family subst_internal.
Final Field tm_true : ...
  := fun x t => tm_true.
(* ... *)
EndFamily. 
Inherits subst.
(* ... *)
EndFamily.
\end{lstlisting}
  \end{minipage}
  \begin{minipage}[t]{0.65\linewidth}
\begin{minted}[fontsize=\footnotesize,escapeinside=@@]{Coq}
Module Type STLC_Interface.
  Parameter (tm : Set).
  Parameter (tm_var : id → tm).
  Parameter (tm_abs : id → tm → tm).
  (* ... and all the fields defined before
  and the abstraction for inductive type ...*)
End STLC_Interface.
(* Collect eliminator to Set for tm *)
Module tm_rec_type (STLC: STLC_Interface).
  Definition __recursor_type_tm_rec :=
    forall P : STLC.tm -> Set,
    (forall n : id, P (STLC.tm_var n)) ->
    (forall (n : id) (i : STLC.tm),
    P i -> P (STLC.tm_abs n i)) ->
    ... -> forall i : STLC.tm, P i.
End tm_rec_type.
(* Compiled Field STLC.subst_internal.tm_var *)
Module tm_var_handle (self__STLC: STLC_Interface)
  (self__subst_internal: EmptySig_84).
Definition tm_var :
  forall (s : ident) (x : ident) (t : self__STLC.tm), self__STLC.tm :=
  fun s x t => if eqb x s then t else self__STLC.tm_var s.
End tm_var_handle. (* ... and more *)
(* Assembling STLC.subst_internal when compilation *)
Module subst_internal_410 (STLC: STLC_Interface). 
  Module subst_internal.
  Include tm_var_handle STLC. Include tm_abs_handle STLC.
  Include tm_app_handle STLC. Include tm_unit_handle STLC.
  End subst_internal.
End subst_internal_410.
Module do_tc (STLC: STLC_Interface).
(* Intermediate Module solely for type checking *)
  Include tm_rec_type STLC.
  Parameter (recursor_for_type_checking : __recursor_type_tm_rec).
  Definition term_for_type_checking :=
    recursor_for_type_checking (fun _  ⇒ id → STLC.tm → STLC.tm)
      subst_internal_410.tm_var
      subst_internal_410.tm_abs ... .
End do_tc.
(* Ctx that expose concrete inductive type *)
Module Type ctx_subst_113.
Inductive __internal_tm : Set := ... 
Include subst_internal_410. (* ... and more inclusion *)
End ctx_subst_113.
(* Compilation of recursor, not reusable
    when inductive type extended due to the ctx *)
Module subst_128 (self__STLC: ctx_subst_113).
Include __motive_of_subst_107 self__STLC.
Definition subst :=
  self__STLC.__internal_tm_rec __motiveTsubst
	self__STLC.subst_internal.tm_var self__STLC.subst_internal.tm_abs
    self__STLC.subst_internal.tm_app self__STLC.subst_internal.tm_unit.
End subst_128.
(* Assembling the whole STLC *)
Module STLC.
Include tm_5. Include tm_prec_72. Include tm_prectm_81.
Include subst_internal_410. Include subst_128.
End v_129.
(* Assembling STLC_bool.subst_internal via reusing 
    most of the former compilation results *)
Module subst_internal_248 (STLC_bool: STLC_Interface).
Include tm_true_197 STLC_bool.
Include tm_false_200 STLC_bool.
Include tm_if_203 STLC_bool.
(* The Reused Compilation result *)
Include tm_var_handle STLC_bool.
Include tm_abs_handle STLC_bool.
Include tm_app_handle STLC_bool.
Include tm_unit_handle STLC_bool.
End v_248.
@\YZ{It'd be useful to (1) show how the parts of the STLC family are assembled and recursor inserted, and (2) show the code compiled for STLC\_bool and how code from STLC is shared with STLC\_bool.}@
\end{minted}
  \end{minipage}
\caption{Exemplar STLC Code, especially about Inductive Type}\label{fig:plugin-example2}
\end{figure}

@@

Let's look at the exemplar \cref{fig:plugin-example2}, the example of
"subst" on STLC. When we define the extensible inductive type "tm", all
the following fields (e.g., "subst_internal.tm_var") can only know "tm"
is a type (of "Set") and there are at least two constructors for it,
just like what we declare in ``the abstracted interface''---the compiled
module type "STLC_Interface". We cannot export
%\YZ{export or put?}\EDJreply{Ok.
%I just figure out how my wording is confusing. put is absolutely correct
%but by put I want to mean ``export''. I think I use ``export'' in
%several places...}
the eliminator of the inductive type "tm" to "STLC_Interface".
Otherwise all the subsequent fields (e.g. "subst_internal.tm_var" and
the corresponding "tm_var_411") would not be able to be inherited to the
context where "tm" is extended with a third constructor, \textbf{since
the mere existence of the recursor will assert that there are only two
constructors} in "STLC_bool", not \textbf{extensible} at all!

However, we later still need the concrete eliminator on "tm" for doing
exhaustiveness checking (as in \ref{chg:extensible-inductive-type}),
thus we will store the corresponding info somewhere.  Here we show one
exemplar extraction of the eliminator to \mintinline{Coq}{Set} as module
"tm_rec_type". This concrete eliminator will be used by our "FRecursor".
For example, in \cref{fig:plugin-example2}, "FRecursor" command will
feed Coq a module "do_tc" for doing type-checking with specified
handlers (e.g. "neg_handler") during compilation---what it does is
basically
assert the existence of the recursor and apply to see if handlers can
pass the type checking by Coq.\YZ{
  Sounds like we can still claim the plugin is doing type checking.
  It's just that the checking is done by generating some Coq code and
  then having Coq check it.
}\EDJreply{Yes! If my phrasing doesn't make things clear (i.e. our plugin is doing typechecking),
feel free to change the phrasing for me}
Note that, at this point of the example,
"FRecursor" will only do type check. The compilation of "FRecursor" into
a component of the compiled module only happens when closing the whole
family, by inserting the following into the compiled module: 
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and
the (vanilla) inductive type in that compiled module. Thus a "FRecursor"
field is unlike other fields, even though it is exposed as a plain
function for the succeeding fields in the programming interface.

@@

Importantly, the recursor field (e.g. "subst") is not actually
inheritable, because it is not compiled like any other fields.
Recall how other fields are compiled into reusable pieces of
(parametrized) modules; a recursor field is not compiled like that.
For all the children family, we have to reconstruct and override the
"FRecursor" field with a ``new'' recursor using (maybe the same) handler
family.  This is affordable since each aggregated recursor handler
doesn't need to be rechecked---look at "STLC_bool.neg_handler", there
are only  new handlers for "tm_true" and the handlers (mundane fields)
from the parent "B.subst_internal.tm_var" have already been type-checked
and inherited (since they only rely on ``the abstracted interface''
module "STLC_Interface"). 
Thanks to family inheritance, their corresponding compiled module from the parents---for example, 
the module "tm_var_411"---doesn't require another type-check and is
re-used during the type-checking and compilation of "STLC_bool.subst".
For example, for the new "subst", when compiling family "STLC_bool" into
a module, at the appropriate place our plugin will insert
\begin{minted}{Coq}
Definition subst := tm_rec (λ _ ⇒ id → tm → tm) subst_internal.tm_var ...
\end{minted}
where "tm_rec" and "tm" are the corresponding recursor and the (vanilla
but extended and modified) inductive type in that compiled module.\YZ{Where is this "tm" inductive type in figure 6?} Note
that, even though "tm_rec" and "tm" has the same name as before, they
are in the newly-compiled module so they refer to totally different
things.

@@

Though partial recursor is provable by simply using "FRecursor", for convenience, we support using "FScheme" to automatic generate it. At the meantime, we implement a tactic "prec_discriminate" that when fed with the appropriate partial recursor (with computational axiom generated), it can discriminate an absurd equation between two different constructors of extensible inductive type.
\begin{minted}{Coq}
  FInductive tm : Set := tunit : tm | tvar : ident -> tm | ..
  FScheme tm_prec PRecT about self__STLC.tm.
  FScheme Rec2D about self__STLC.tm_prec.
  (* ... *)
  Goal ∀ i, tvar i = tunit -> False.
  intros i H. prec_discriminate self__STLC.tm_prec H. Qed.
\end{minted}
